---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
# Reproducible Research: Peer Assessment 1

Please note for this code to work you must have downloaded the plyr, ggplot, grid, and scales packages.  This program will run the library statements for those packages.  
Also note that this program uses the current working directory. 

# Loading and preprocessing the data
### Make sure libraries are installed
```{r ,echo=TRUE}
library(plyr)
library(ggplot2)
library(scales)
library(grid)
```

### Import and process the data to adjust data types and check the number of rows in the data to compare to documentation provided
```{r ,echo=TRUE}
setInternet2(use = TRUE)
fileurl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileurl,destfile="activity.zip")
rawactivity <- read.csv(unz("activity.zip","activity.csv"),header=TRUE)

nrow(rawactivity)

rawactivity["newdate"] <- as.Date(rawactivity$date)
```
Notice that the interval field is not in minutes but in hours:minutes example goes from 55 to 100 instead of 55 to 60
```{r ,echo=TRUE}
rawactivity["newinterval"]<-  paste0(substr(sprintf("%04d",rawactivity$interval),1,2),":",
                                     substr(sprintf("%04d",rawactivity$interval),3,4),":00")
rawactivity["intervaldatetime"]<-paste(rawactivity$date,rawactivity$newinterval)
rawactivity["stepsdatetime"]<-as.POSIXct(rawactivity$intervaldatetime)

activity <- rawactivity[c("steps","stepsdatetime")]
str(activity)
```


# What is the mean total number of steps taken per day?

### What is the distribution of total steps taken per day and the mean and median of steps per day
```{r ,echo=TRUE}
DailyStepTotal  <- aggregate(activity$steps,by=list(as.factor(format(activity$stepsdatetime, "%Y%m%d"))),
                             FUN=sum)
DailyStepTotal = rename(DailyStepTotal,c("Group.1"="Day","x"="TotalSteps"))
DailyStepMean   <- aggregate(activity$steps,by=list(as.factor(format(activity$stepsdatetime, "%Y%m%d"))),
                             FUN=mean)
DailyStepMean = rename(DailyStepMean,c("Group.1"="Day","x"="MeanSteps"))
DailyStepMedian <- aggregate(activity$steps,by=list(as.factor(format(activity$stepsdatetime, "%Y%m%d"))),
                             FUN=median)
DailyStepMedian = rename(DailyStepMedian,c("Group.1"="Day","x"="MedianSteps"))

hist(DailyStepTotal$TotalSteps, breaks = 20, xlab = "Steps", main = "Histogram of Total Steps Per Day")

DailyTotals  <- merge(DailyStepMean,DailyStepMedian,by="Day")
print(DailyTotals, type = "html")
```


# What is the average daily activity pattern?
### What is the average steps per interval averaged across all of the days in the data and what time of day had the most average steps
### A time series plot over all 288 intervals reveals some interesting patterns in the data.
```{r ,echo=TRUE}
IntervalStepMean   <- aggregate(activity$steps,by=list(as.factor(format(activity$stepsdatetime, "%H%M"))),
                                FUN=mean,na.rm = TRUE)
IntervalStepMean = rename(IntervalStepMean,c("Group.1"="Interval","x"="MeanSteps"))
ggplot(data = IntervalStepMean, aes(x=Interval,y = MeanSteps,group=1)) + 
        geom_line()+ theme_bw()+ labs(x="Time of Day") +
        scale_x_discrete(breaks = c("0000","0400","0800","1200","1600","2000","2355")) +
        labs(y = expression("Average Number of Steps"))+ ggtitle("Average Number of Steps During a Day")+
        theme(plot.title = element_text(face="bold"))
MaxSteps <- max(IntervalStepMean$MeanSteps)
MaxStepsInterval = IntervalStepMean[IntervalStepMean$MeanSteps == MaxSteps,]

MaxStepsTime <- paste0(substr(MaxStepsInterval$Interval,1,2),":",
                       substr(MaxStepsInterval$Interval,3,4))
MaxStepsfrmt <- format(max(IntervalStepMean$MeanSteps),scientific=FALSE,digits=3)

```
### The interval with the most average steps was `r MaxStepsTime` with an average of `r MaxStepsfrmt`.

# Imputing missing values
The technique used here was to apply the mean for the interval to those intervals with a missing value of NA.

Determine how many rows have NA values for the step column
```{r ,echo=TRUE}
narowscount = format(nrow(activity[is.na(activity$steps),]),big.mark=",",scientific=FALSE)
totalrecordcount = format(nrow(activity),big.mark=",",scientific=FALSE)
nadayscount = format(nrow(DailyStepTotal[is.na(DailyStepTotal$TotalSteps),]),big.mark=",",scientific=FALSE)
totaldaycount = format(nrow(DailyStepTotal),big.mark=",",scientific=FALSE)
```
### There were `r narowscount` rows that had NA values out of `r totalrecordcount` rows.  
### Out of `r totaldaycount` days there were `r nadayscount` days for which all of the data was missing. 

Impute the NA values by merging the interval mean back to the original dataframe and assigning  
the mean value to any steps for wich the value was NA.  
Essentially NA step values are substituted with the mean for that interval across all days in the dataframe.  
```{r ,echo=TRUE}
activity["Interval"] <- format(activity$stepsdatetime, "%H%M")
ActivityImputed  <- merge(IntervalStepMean,activity,by="Interval")
ActivityImputed["ImputedSteps"] <- ifelse(is.na(ActivityImputed$steps),ActivityImputed$MeanSteps,ActivityImputed$steps)
```

### What is the distribution of total steps taken per day and the mean and median of steps per day  
### based on the newly imputed steps variable
```{r ,echo=TRUE}
DailyStepTotalIMP  <- aggregate(ActivityImputed$ImputedSteps,by=list(as.factor(format(ActivityImputed$stepsdatetime, "%Y%m%d"))),
                                FUN=sum)
DailyStepTotalIMP = rename(DailyStepTotalIMP,c("Group.1"="Day","x"="TotalImputedSteps"))
DailyStepMeanIMP   <- aggregate(ActivityImputed$ImputedSteps,by=list(as.factor(format(ActivityImputed$stepsdatetime, "%Y%m%d"))),
                                FUN=mean)
DailyStepMeanIMP = rename(DailyStepMeanIMP,c("Group.1"="Day","x"="MeanImputedSteps"))
DailyStepMedianIMP <- aggregate(ActivityImputed$ImputedSteps,by=list(as.factor(format(ActivityImputed$stepsdatetime, "%Y%m%d"))),
                                FUN=median)
DailyStepMedianIMP = rename(DailyStepMedianIMP,c("Group.1"="Day","x"="MedianImputedSteps"))

hist(DailyStepTotalIMP$TotalImputedSteps, breaks = 20, xlab = "Steps", main = "Histogram of Total Steps Per Day Imputed")

##Merge the mean and median file created above to create a table with both and write that table out
DailyTotalsIMP  <- merge(DailyStepMeanIMP,DailyStepMedianIMP,by="Day")
print(DailyTotalsIMP, type = "html")

DailyTotalsBoth <- merge(DailyTotals,DailyTotalsIMP,by="Day")
print(DailyTotalsBoth, type = "html")
```
There is no difference in each days mean and median unless that day was NA before.  
The reason for this is that the day was 100% populated or 100% NA.  
For those days that were 100% NA my technique was to substitute the mean for each interval.  
The result was that for days that were previously NA the new total steps are 10,766.  
The mean for those days is 37.383 and median of 34.113.  
Becuase of the 8 new days with 10,766 steps the histogram did larger concenration in that range.  

# Are there differences in activity patterns between weekdays and weekends?
Based on the graph below it appears that the activity is more consistent through the day on weekends.  
However, the most steps on average for a given time of day still occur arount 8:00AM in the morning on weekdays.  
```{r ,echo=TRUE}
##does activity on weekends differ from activity during the week
##create a weekend variable by checking to see if the day of the week starts with an S (Saturday or Sunday)
ActivityImputed["Weekend"] <- as.factor(ifelse(substr(weekdays(as.POSIXct(ActivityImputed$stepsdatetime),
                                                               abbreviate=TRUE),1,1)=="S","weekend","weekday"))
WEEKEND <- ActivityImputed[ActivityImputed$Weekend == "weekend",]
WEEKDAY <- ActivityImputed[ActivityImputed$Weekend == "weekday",]

WEEKENDSTEPMEAN   <- aggregate(WEEKEND$ImputedSteps,by=list(as.factor(format(WEEKEND$stepsdatetime, "%H%M"))),
                               FUN=mean)
WEEKENDSTEPMEAN = rename(WEEKENDSTEPMEAN,c("Group.1"="Interval","x"="MeanSteps"))
WEEKDAYSTEPMEAN   <- aggregate(WEEKDAY$ImputedSteps,by=list(as.factor(format(WEEKDAY$stepsdatetime, "%H%M"))),
                               FUN=mean)
WEEKDAYSTEPMEAN = rename(WEEKDAYSTEPMEAN,c("Group.1"="Interval","x"="MeanSteps"))

##THIS FUNCTION WAS COPIED FROM THE URL http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
## The cookbook for R 
multiplot <- function(..., plotlist=NULL, cols) {
        require(grid)
        
        # Make a list from the ... arguments and plotlist
        plots <- c(list(...), plotlist)
        
        numPlots = length(plots)
        
        # Make the panel
        plotCols = cols                          # Number of columns of plots
        plotRows = ceiling(numPlots/plotCols) # Number of rows needed, calculated from # of cols
        
        # Set up the page
        grid.newpage()
        pushViewport(viewport(layout = grid.layout(plotRows, plotCols)))
        vplayout <- function(x, y)
                viewport(layout.pos.row = x, layout.pos.col = y)
        
        # Make each plot, in the correct location
        for (i in 1:numPlots) {
                curRow = ceiling(i/plotCols)
                curCol = (i-1) %% plotCols + 1
                print(plots[[i]], vp = vplayout(curRow, curCol ))
        }
        
}

p1 <-
ggplot(data = WEEKENDSTEPMEAN, aes(x=Interval,y = MeanSteps,group=1)) + 
        geom_line()+ theme_bw()+ labs(x="Time of Day") +
        scale_x_discrete(breaks = c("0000","0400","0800","1200","1600","2000","2355")) +
        labs(y = expression("Average Number of Steps"))+ ggtitle("Average Number of Steps During a Weekend")+
        ylim(0,250) +
        theme(plot.title = element_text(face="bold"))
p2 <-
ggplot(data = WEEKDAYSTEPMEAN, aes(x=Interval,y = MeanSteps,group=1)) + 
        geom_line()+ theme_bw()+ labs(x="Time of Day") +
        scale_x_discrete(breaks = c("0000","0400","0800","1200","1600","2000","2355")) +
        labs(y = expression("Average Number of Steps"))+ ggtitle("Average Number of Steps During a Weekday")+
        ylim(0,250) +
        theme(plot.title = element_text(face="bold"))
multiplot(p1, p2, cols=1)
```

